Forcing for Isabelle

### 2018-02-12 blueprint.tex
intro to forcing. Arriving to hard stuff

\documentclass[11pt,english]{article}
\usepackage[bottom=2cm, top=3cm, %% Comment if 
  left=4cm, right=4cm]{geometry}             %% using 'ebook' style

\input{header-blueprint}
%%%%%%%%%%%%%%%%%%% % %   Unicodings   % % %%%%%%%%%%%%%%%%%%%%%
\newcommand{\mt}[1]{$\mathtt{#1}$}
\DeclareUnicodeCharacter{21D2}{\mt\Rightarrow}
\DeclareUnicodeCharacter{2208}{\mt\in}
\DeclareUnicodeCharacter{2200}{\mt\forall}
\DeclareUnicodeCharacter{27F6}{\mt\longrightarrow}
\DeclareUnicodeCharacter{27E6}{\mt\llbracket}
\DeclareUnicodeCharacter{27E7}{\mt\rrbracket}
\DeclareUnicodeCharacter{27F9}{\mt\Longrightarrow}
\DeclareUnicodeCharacter{03A6}{\mt\Phi}
%\DeclareUnicodeCharacter{}{\mt}

\renewcommand{\PP}{\mathbb{P}}
\newcommand{\formula}{\ensuremath{\mathtt{formula}}}
\newcommand{\tyi}{\mathtt{i}}
\newcommand{\tyo}{\mathtt{o}}
\DeclareMathOperator{\val}{val}

@@

\section{Names}
The following is mostly independent of the previous section, but we
assume that $\1\in\PP\in M$, $M$ is transitive and $1\in G\sbq\PP$.
(Note that, in general, $G\notin M$.)

@@

Given a ctm $M$, and a $M$-generic filter $G\sbq\PP$, the Forcing
Theorems relate satisfaction of a formula 
$\phi$ in the generic extension $M[G]$ to the satisfaction of another formula
$\phi'$ in $M$. The map $\phi\mapsto\phi'$ is defined by recursion on
the structure of $\phi$.

>>>>

\documentclass[11pt,english]{article}
\usepackage[bottom=2cm, top=3cm, %% Comment if 
  left=4cm, right=4cm]{geometry}             %% using 'ebook' style 
\usepackage{framed}

\input{header-blueprint}
%%%%%%%%%%%%%%%%%%% % %   Unicodings   % % %%%%%%%%%%%%%%%%%%%%%
\newcommand{\mt}[1]{$\mathtt{#1}$}
\DeclareUnicodeCharacter{21D2}{\mt\Rightarrow}
\DeclareUnicodeCharacter{2208}{\mt\in}
\DeclareUnicodeCharacter{2200}{\mt\forall}
\DeclareUnicodeCharacter{27F6}{\mt\longrightarrow}
\DeclareUnicodeCharacter{27E6}{\mt\llbracket}
\DeclareUnicodeCharacter{27E7}{\mt\rrbracket}
\DeclareUnicodeCharacter{27F9}{\mt\Longrightarrow}
\DeclareUnicodeCharacter{03A6}{\mt\Phi}
\DeclareUnicodeCharacter{03BB}{\mt\lambda}
%\DeclareUnicodeCharacter{}{\mt}

\renewcommand{\PP}{\mathbb{P}}
\newcommand{\formula}{\ensuremath{\mathtt{formula}}}
\newcommand{\tyi}{\mathtt{i}}
\newcommand{\tyo}{\mathtt{o}}
\newcommand{\forceisa}{\mathtt{forces}}
\DeclareMathOperator{\val}{val}

@@

\section{Names}
\begin{framed}
  The following is mostly independent of the previous section, but 
  \begin{bfseries}
    we assume that $\1\in\PP\in M$, $M$ is transitive and $1\in
    G\sbq\PP$.
  \end{bfseries}
  (Note that, in general, $G\notin M$.)
\end{framed}

@@

\begin{framed}
  It might be necessary to redefine \verb|sats| and the related
  function \verb|satisfies| in order to have an independent argument $B$
  to be the range of values of the environment. We would have, for
  example, 
\begin{verbatim}
consts   satisfies :: "[i,i]=>i"
primrec 
  "satisfies(A,B,Member(x,y)) =
      (λenv ∈ list(B). bool_of_o (nth(x,env) ∈ nth(y,env)))"
\end{verbatim}
  \dots

\noindent and then
\begin{verbatim}
abbreviation
  sats :: "[i,i,i] => o" where
  "sats(A,p,env) == satisfies(A,A,p)`env = 1"
\end{verbatim}
\end{framed}

Given a ctm $M$, and an $M$-generic filter $G\sbq\PP$, the Forcing
Theorems relate satisfaction of a formula 
$\phi$ in the generic extension $M[G]$ to the satisfaction of another formula
$\phi'$ in $M$. The map $\phi\mapsto\phi'$ is defined by recursion on
the structure of $\phi$. The fact that this map works (in particular,
in order to show that $M[G]\models\ZFC$), we must use the fact that
$M\models\ZFC$. Since we can only assert the latter using internalized
formulas, the aformentioned map will be defined as a function from the
set \formula{} into itself.

Up to this point, the main reason for working with internalized
versions is that  it is not possible to do recursion with formulas of
type $\tyo$ (i.e., the formulas in the first-order logic of
Isabelle/ZF).

We will now make more precise the definition of the map
$\phi\mapsto\phi'$ and how it relates satisfaction in $M$ to that in
$M[G]$. Actually, if the formula $\phi$ has $n$ free variables,
$\phi'$ will have $n+4$ free variables, where the first four account
for the forcing notion and a particular condition. If
$\phi=\phi(x_1,\dots,x_n)$, the standard notation for $\phi'$ is
\[
p\forces_{\PP,\leq,\1}^* \phi(\tau_1,\dots,\tau_n).
\]
Here, $\PP,\leq,\1,$ and $p\in\PP$ are the extra parameters. For the
time being, we just consider $\tau_1,\dots,\tau_n$ to be ordinary
variables. Informally, we call $\forces$ the \emph{forcing relation}.
Given $\quine\phi\in\formula$, we will write the forcing relation in
Isabelle/ZF by using  
\[
\forceisa(\quine\phi,\PP,\leq,\1)
\]

The base case (of atomic formulas) is famously the most
complicated one.
